New lower bounds for Schur and weak Schur numbers

### 2021-05-09 article/schur.tex
schur p to q n to k

\begin{theorem}
	Let \((p,k), (q,n) \in (\mathbb{N}^*)^2\). If there exists a \(SF\)-template with width \(p\) and \(k+1\) colors 
	and a partition of \([\![1,q]\!]\) into \(n\) sum-free subsets then there exists a partition of \([\![1,pq+m_{k+1}
	-1]\!]\) into \(n+k\) sum-free subsets. \(m_{k+1}\) is the minimum of the special subset in the \(SF\)-template.
\end{theorem}

Setting \(p = S^+(k+1)\) and \(q = S(n)\) yields the following corollary.

\begin{corollary}
	Let \(n, k \in \mathbb{N}^*\). Then
	\[ S(n+k) \geqslant S^+(k+1)S(n) + m_{k+1} - 1 \]
\end{corollary}

@@

\begin{proof}[\textsc{Proof.}]
Denote by \(f\) the coloring associated to the \(SF\)-template with width \(p\) and \(g\) the one associated 
to the sum-free partition of \([\![1,q]\!]\); where \(f : [\![1,p]\!] \longrightarrow [\![1,k+1]\!]\) and 
\(g : [\![1,q]\!] \longrightarrow [\![1,n]\!]\).

\par
NB: In the following 3 predicates, the conditions \(x + y \leqslant p\)  and \(x + y \leqslant q\) are omitted for readability.
\par
The sum-free condition is expressed as
\[\forall (x,y) \in [\![1,p]\!]^2, f(x) = f(y) \Longrightarrow f(x+y) \neq f(x)\]
\[\forall (x,y) \in [\![1,q]\!]^2, g(x) = g(y) \Longrightarrow g(x+y) \neq g(x)\]

The additionnal constraint for the SF-template is
\[
\forall (x,y) \in [\![1,p]\!]^2, \left\{
\begin{array}{l}
	f(x) = f(y) \leqslant k \\
	x + y > p
\end{array}
\right. \Longrightarrow f(x+y-p) \neq f(x)
\]

For \(x \in [\![1,pq+m_{k+1}-1]\!]\), write \(x = (\alpha - 1) + u\) for some integers \(\alpha \in \mathbb{Z}\) and 
\(u \in [\![1,p]\!]\). This decomposition is of course unique. \(\alpha\) (resp. \(u\)) can be interpreted as the row 
(resp. column) number of \(x\). Define a new coloring \(h\) as follows
\[
\begin{array}{c c c l}
	h : & [\![1,pq+m_{k+1}-1]\!] & \longrightarrow & [\![1,n+k]\!] \\
	& x & \longmapsto & 
	\left\{ \begin{array}{l l}
		f(u) & \text{if}~f(u) \leqslant k \\
		k + g(\alpha) & \text{if}~f(u) = k + 1 \\
	\end{array} \right.
\end{array}
\]

\(h\) is well-defined since, by definition of \(m_{k+1}\), \(\forall x \in [\![p q + 1, p q + m_{k + 1} - 1 ]\!], f(u) 
\leqslant k\) and therefore \(\forall x \in [\![1,pq+m_{k+1}-1]\!], f(u) = k + 1 \implies \alpha \in [\![1, q]\!]\).

\par
We now prove that \(h\) is a sum-free coloring. Let \(x,y \in [\![1,pq + m_{k+1}-1]\!]\) such that \(h(x) = h(y)\) 
and \(x+y \leqslant pq+m_{k+1}-1\). We claim that \(h(x+y) \neq h(x)\). Write \(x = (\alpha - 1) p + u\) and 
\(y = (\beta - 1) p + v\) where \(\alpha, \beta \in \mathbb{Z}\) and \(u, v \in [\![1,p]\!]\). Distinguish between 
two cases according to the value of \(h(x)\). \\

@@

\begin{itemize}
\item If \(u + v > p\), write \(w = u + v - p \in [\![1, p]\!]\). Then \(x + y = (\alpha + \beta - 1) p + w\). By definition, 
	\(h(x + y) = f(w)\). Given that \(f(u) = f(v) \leqslant k\), the additionnal constraint on \(f\) implies \(f(w) 
	\neq f(u)\), that is \(h(x + y) \neq h(x)\).
\item If \(u + v \leqslant p\), write \(w = u + v \in [\![1, p]\!]\). Then \(x+y = (\alpha + \beta- 2) p + w\). By definition, 
	\(h(x + y) = f(w)\). Given that \(f(u) = f(v) \leqslant k\), the sum-free property of \(f\) implies \(f(w) \neq f(u)\), 
	that is \(h(x + y) \neq h(x)\).
\end{itemize}

\noindent \underline{\textbf{Case 2:} \(h(x) \geqslant k + 1\)}
\par
\(h(x) = k + g(\alpha) = 
k + g(\beta) = h(y)\), hence \(g(\alpha) = g(\beta)\). As in the first case, distinguish between two cases according 
to the value of \(x + y\).

\begin{itemize}
\item \begin{sloppypar}
	If \(u + v > p\), write \(w = u + v - p \in [\![1, p]\!]\). Then \(x + y = (\alpha + \beta - 1) p + w\). Assume that 
	\({h(x+y) \geqslant k + 1}\),  otherwise \(h(x + y) \neq h(x)\) obviously holds. By definition, \({h(x + y) = k + g(\alpha + 
	\beta)}\). Given that \(g(\alpha) = g(\beta)\), the sum-free property of \(g\) implies \(g(\alpha + \beta) 
	\neq g(\alpha)\) that is \(h(x + y) \neq h(x)\).
	\end{sloppypar}
\item  If \(u + v \leqslant p\), write \(w = u + v \in [\![1, p]\!]\). Then \(x+y = (\alpha + \beta- 2) p + w\). The sum-free 
	property of \(f\) implies \(f(w) \neq f(u)\). Therefore \(f(w) \leqslant k\) and thus \(h(x + y) \leqslant k\). In particular,
	given that \(h(x) \geqslant k + 1\), \(h(x + y) \neq h(x)\).
\end{itemize}
\end{proof}

@@

\begin{proposition}
Let \((p, k) \in \mathbb{N}^*)^2\) and let \(f\) be a coloring associated to a SF-template with width \(p\) and \(k+1\) 
colors. Let \(b \in \mathbb{N}\) (\(b = m_{k+1} - 1\) works) and assume there there exists a coloring \(g\) of 
\([\![1, b]\!]\) with \(k+1\) colors such that

\begin{itemize}
	\item \(\forall (x, y) \in [\![1, p]\!]^2,(f(x) = f(y) \text{ and } (x + y) \mod p \leqslant b) 
	\implies g((x + y) \mod p) \neq f(x)\)
	\item \(\forall (x, y) \in [\![1, p]\!] \times  [\![1, b]\!], (f(x) = g(y) \text{ and } x + y \leqslant b) 
	\implies g(x + y) \neq f(x)\)
\end{itemize}

Then, for every \(n \in \mathbb{N}^*\), by using on the last row the coloring \(i \longmapsto g(i - p S(n))\), we have
\[ S(n+k) \geqslant S^+(k+1)S(n) + b\]
\end{proposition}

@@

\begin{theorem}
Let \((p,k), (q,n) \in (\mathbb{N}^*)^2\). If there exists a SF-template with width \(p\) and \(k+1\) colors,
and a SF-template with width \(q\) and \(n\) colors, then there exists SF-template with width \(pq\) and \((n+k)\) 
colors.
\end{theorem}

@@

\begin{corollary}
Let \(n, k \in \mathbb{N}^*\). Then
\[ S^+(n+k) \geqslant S^+(k+1)S^+(n) \]
\end{corollary}

>>>>

\begin{theorem}
	Let \((p,k), (q,n) \in (\mathbb{N}^*)^2\). If there exists a \(SF\)-template with width \(q\) and \(n+1\) colors 
	and a partition of \([\![1,p]\!]\) into \(k\) sum-free subsets then there exists a partition of \([\![1,pq+m_{n+1}
	-1]\!]\) into \(n+k\) sum-free subsets. \(m_{n+1}\) is the minimum of the special subset in the \(SF\)-template.
\end{theorem}

Setting \(q = S^+(n+1)\) and \(p = S(k)\) yields the following corollary.

\begin{corollary}
	Let \(n, k \in \mathbb{N}^*\). Then
	\[ S(n+k) \geqslant S^+(n+1)S(k) + m_{n+1} - 1 \]
\end{corollary}

@@

\begin{proof}[\textsc{Proof.}]
	Denote by \(f\) the coloring associated to the \(SF\)-template with width \(q\) and \(g\) the one associated 
	to the sum-free partition of \([\![1,p]\!]\); where \(f : [\![1,q]\!] \longrightarrow [\![1,n+1]\!]\) and 
	\(g : [\![1,p]\!] \longrightarrow [\![1,k]\!]\).

	\par
	NB: In the following 3 predicates, the conditions \(x + y \leqslant q\)  and \(x + y \leqslant p\) are omitted for readability.
	\par
	The sum-free condition is expressed as
	\[\forall (x,y) \in [\![1,q]\!]^2, f(x) = f(y) \Longrightarrow f(x+y) \neq f(x)\]
	\[\forall (x,y) \in [\![1,p]\!]^2, g(x) = g(y) \Longrightarrow g(x+y) \neq g(x)\]

	The additionnal constraint for the SF-template is
	\[
	\forall (x,y) \in [\![1,q]\!]^2, \left\{
	\begin{array}{l}
		f(x) = f(y) \leqslant n \\
		x + y > q
	\end{array}
	\right. \Longrightarrow f(x+y-q) \neq f(x)
	\]

	For \(x \in [\![1,pq+m_{n+1}-1]\!]\), write \(x = (\alpha - 1) + u\) for some integers \(\alpha \in \mathbb{Z}\) and 
	\(u \in [\![1,q]\!]\). This decomposition is of course unique. \(\alpha\) (resp. \(u\)) can be interpreted as the row 
	(resp. column) number of \(x\). Define a new coloring \(h\) as follows
	\[
	\begin{array}{c c c l}
		h : & [\![1,pq+m_{n+1}-1]\!] & \longrightarrow & [\![1,n+k]\!] \\
		& x & \longmapsto & 
		\left\{ \begin{array}{l l}
			f(u) & \text{if}~f(u) \leqslant n \\
			k + g(\alpha) & \text{if}~f(u) = n + 1 \\
		\end{array} \right.
	\end{array}
	\]

	\(h\) is well-defined since, by definition of \(m_{n+1}\), \(\forall x \in [\![p q + 1, p q + m_{n + 1} - 1 ]\!], f(u) 
	\leqslant k\) and therefore \(\forall x \in [\![1,pq+m_{n+1}-1]\!], f(u) = n + 1 \implies \alpha \in [\![1, p]\!]\).

	\par
	We now prove that \(h\) is a sum-free coloring. Let \(x,y \in [\![1,pq + m_{n+1}-1]\!]\) such that \(h(x) = h(y)\) 
	and \(x+y \leqslant pq+m_{n+1}-1\). We claim that \(h(x+y) \neq h(x)\). Write \(x = (\alpha - 1) q + u\) and 
	\(y = (\beta - 1) q + v\) where \(\alpha, \beta \in \mathbb{Z}\) and \(u, v \in [\![1,q]\!]\). Distinguish between 
	two cases according to the value of \(h(x)\). \\

@@

	\begin{itemize}
	\item If \(u + v > q\), write \(w = u + v - q \in [\![1, q]\!]\). Then \(x + y = (\alpha + \beta - 1) q + w\). By definition, 
		\(h(x + y) = f(w)\). Given that \(f(u) = f(v) \leqslant n\), the additionnal constraint on \(f\) implies \(f(w) 
		\neq f(u)\), that is \(h(x + y) \neq h(x)\).
	\item If \(u + v \leqslant p\), write \(w = u + v \in [\![1, q]\!]\). Then \(x+y = (\alpha + \beta- 2) q + w\). By definition, 
		\(h(x + y) = f(w)\). Given that \(f(u) = f(v) \leqslant n\), the sum-free property of \(f\) implies \(f(w) \neq f(u)\), 
		that is \(h(x + y) \neq h(x)\).
	\end{itemize}

	\noindent \underline{\textbf{Case 2:} \(h(x) \geqslant n + 1\)}
	\par
	\(h(x) = n + g(\alpha) = 
	k + g(\beta) = h(y)\), hence \(g(\alpha) = g(\beta)\). As in the first case, distinguish between two cases according 
	to the value of \(x + y\).

	\begin{itemize}
	\item \begin{sloppypar}
		If \(u + v > q\), write \(w = u + v - q \in [\![1, q]\!]\). Then \(x + y = (\alpha + \beta - 1) q + w\). Assume that 
		\({h(x+y) \geqslant n + 1}\),  otherwise \(h(x + y) \neq h(x)\) obviously holds. By definition, \({h(x + y) = n + g(\alpha + 
		\beta)}\). Given that \(g(\alpha) = g(\beta)\), the sum-free property of \(g\) implies \(g(\alpha + \beta) 
		\neq g(\alpha)\) that is \(h(x + y) \neq h(x)\).
		\end{sloppypar}
	\item  If \(u + v \leqslant q\), write \(w = u + v \in [\![1, q]\!]\). Then \(x+y = (\alpha + \beta- 2) q + w\). The sum-free 
		property of \(f\) implies \(f(w) \neq f(u)\). Therefore \(f(w) \leqslant k\) and thus \(h(x + y) \leqslant n\). In particular,
		given that \(h(x) \geqslant n + 1\), \(h(x + y) \neq h(x)\).
	\end{itemize}
\end{proof}

@@

\begin{proposition}
	Let \((q, n) \in \mathbb{N}^*)^2\) and let \(f\) be a coloring associated to a SF-template with width \(q\) and \(n+1\) 
	colors. Let \(b \in \mathbb{N}\) (\(b = m_{n+1} - 1\) works) and assume there there exists a coloring \(g\) of 
	\([\![1, b]\!]\) with \(n+1\) colors such that

	\begin{itemize}
		\item \(\forall (x, y) \in [\![1, q]\!]^2,(f(x) = f(y) \text{ and } (x + y) \mod q \leqslant b) 
		\implies g((x + y) \mod q) \neq f(x)\)
		\item \(\forall (x, y) \in [\![1, q]\!] \times  [\![1, b]\!], (f(x) = g(y) \text{ and } x + y \leqslant b) 
		\implies g(x + y) \neq f(x)\)
	\end{itemize}

	Then, for every \(n \in \mathbb{N}^*\), by using on the last row the coloring \(x \longmapsto g(x - p S(n))\), we have
	\[ S(n+k) \geqslant S^+(n+1)S(k) + b\]
\end{proposition}

@@

\begin{theorem}
	Let \((p,k), (q,n) \in (\mathbb{N}^*)^2\). If there exists a SF-template with width \(q\) and \(n+1\) colors,
	and a SF-template with width \(p\) and \(k\) colors, then there exists SF-template with width \(pq\) and \((n+k)\) 
	colors.
	\end{theorem}

@@

\begin{corollary}
	Let \(n, k \in \mathbb{N}^*\). Then
	\[ S^+(n+k) \geqslant S^+(n+1)S^+(k) \]
\end{corollary}
